---
title: "20180206 NSF meeting"
author: "Miao Cai^[Department of Epidemiology and Biostatistics, College for Public Health and Social Justice, Saint Louis University. Email: [miao.cai@slu.edu](mailto:miao.cai@slu.edu)]"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    number_sections: true
  html_document: default
link-citations: true
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
```

```{r eval=FALSE}
pacman::p_load(tidyverse, broom, rstanarm, brms, bbplot)

sub_cumdrive_slope = function(path, offset = FALSE){
  quanfiles = gsub(".rds", "", list.files(path))
  quanpath = paste0(path, list.files(path))
  
  quanfit = list()
  quanread = function(filename, path){
    quanfit[[filename]] = readRDS(path)
  }
  
  for (i in seq_along(quanfiles)) {
    temt = readRDS(quanpath[i])
    quanfit[[quanfiles[i]]] = broom::tidy(temt, intervals = TRUE, prob = 0.95)
  }
  
  quantable = bind_rows(quanfit, .id = "column_label")
  if(offset){
    quantable$column_label = as.numeric(gsub(".*([0-9]{2})_.*", "\\1", quantable$column_label))
  }else{
    z1 = as.numeric(substr(quantable$column_label, 7, 8))
    z1[is.na(z1)] = (as.numeric(substr(quantable$column_label, 10, 11)) - 1)[is.na(z1)]
    quantable$column_label = z1
    rm(z1)
  }
  
  return(quantable)
}

logitsub = sub_cumdrive_slope("data/CEratequantile/")
cpoi5sub = sub_cumdrive_slope("data/CEratequantileCPOIS/")
cpoi4sub = sub_cumdrive_slope("data/CPOIS4/")
cpoi3sub = sub_cumdrive_slope("data/CPOIS3/")
cpoi2sub = sub_cumdrive_slope("data/CPOIS2/")
logitOFFSETsub = sub_cumdrive_slope("data/cpoisnooffset/")
cpoisOFFSETsub = sub_cumdrive_slope("data/LogitNOoffset/", TRUE)

saveRDS(quantable, "data/logitsub.rds")
saveRDS(cpoi5sub, "data/cpoi5sub.rds")
saveRDS(cpoi4sub, "data/cpoi4sub.rds")
saveRDS(cpoi3sub, "data/cpoi3sub.rds")
saveRDS(cpoi2sub, "data/cpoi2sub.rds")
saveRDS(logitOFFSETsub, "data/logitOFFSETsub.rds")
saveRDS(cpoisOFFSETsub, "data/cpoisOFFSETsub.rds")
```

```{r}
pacman::p_load(tidyverse)
rds_vec = c("logitsub", "cpoi5sub", "cpoi4sub", "cpoi3sub", "cpoi2sub", "logitOFFSETsub", "cpoisOFFSETsub")

for (i in rds_vec) {
  assign(i, readRDS(paste0("data/", i, ".rds")))
}

plot_cumdrive_slp = function(dat, plot_title){
  p = get(dat) %>% 
    filter(term %in% c("cumDrive", "b_cumDrive")) %>% 
    ggplot(aes(x = column_label, y = estimate)) + 
    geom_point(color="red") + geom_errorbar(aes(ymax = upper, ymin = lower)) +
    geom_hline(yintercept = 0, color = "blue", size=2,alpha=0.5) + theme_bw() + 
    ggtitle(plot_title) + xlab("quantiles of driver's CT rates (%)")
}

for (i in rds_vec) {
  assign(i, plot_cumdrive_slp(dat = i, plot_title = i))
}
```

# At what stage did the drivers have critival events?

```{r}
t = readRDS("data/t.Rds")

pacman::p_load(tidyverse)

dat = t %>% 
  select(driver1, driver2, shift, begDate, EndTime, 
         travelTime, waiting_time,
         cumDrive, cumTravel, distance, cnum) %>% 
  mutate(travelTime = round(travelTime/60, 3),
         waiting_time = round(waiting_time/60, 3))

dri_shift = dat %>% group_by(driver1, shift) %>% 
  summarise(shift_time = max(cumDrive))

dat = dat %>% 
  left_join(dri_shift, by = c("driver1", "shift"))

z = dat %>% 
  filter(cnum>0) %>% 
  mutate(CTpercent = cumDrive/shift_time) %>% 
  arrange(desc(CTpercent)) %>% 
  mutate(plotid = 1:n()) 


 z %>% 
  ggplot(aes(plotid, CTpercent)) + geom_point(alpha=0.3,size=0.02) + 
  geom_hline(yintercept = 0.5, color = "blue", size=2,alpha=0.5) + theme_bw()+ 
  xlab("Index of each CE-trip") + 
  ylab("Cumulative driving time over total shift time")
```

\newpage

# Subgroup analysis
## Logistic regression for subgroups

- Calculate the rate of critical events for each driver
- Sort the rate of critical events
- Divide the drivers into 20 subgroups according to the quantiles of CT rates
- Construct random intercepts and random slopes models for these subgroups

```{r}
logitsub
```

\newpage

## After removing the length of the trip - logit

```{r}
logitOFFSETsub
```


\newpage

## Censored Poisson regression for subgroups

I also tested the sensitivity of the threshold 5 for censored Poisson regression.

```{r gridcpois}
pacman::p_load(gridExtra)
grid.arrange(cpoi5sub, cpoi4sub, cpoi3sub, cpoi2sub)
```

## After removing the length of the trip - censored Poisson

```{r}
cpoisOFFSETsub
```


\newpage

# Checking the association between trip time and cumulative driving time

## Boxplot on the distribution of trip time between CT and non-CT trips
```{r}
t = readRDS("data/t.Rds")

t %>% 
  mutate(CT_binary = ifelse(cnum > 0, "Yes", "No"),
         travelTime = round(travelTime/60, 2)) %>% 
  ggplot(aes(CT_binary, travelTime)) +
  geom_violin(alpha = 0.3, width = 1.5, size = 1, fill = "#1F3552", color = "#1F3552") +
  geom_boxplot(width = 0.5, alpha = 0.8) + theme_bw() + 
  xlab("Whether critical event occurred or not") + ylab("length of the trip (hours)") + 
  ggtitle("Boxplot of trip time by CT")+ coord_flip()
# geom_jitter(width = 0.2, alpha = 0.01)
```

\newpage

## Scatterplot on trip time and cumulative driving time
```{r}
t %>% 
  mutate(CT_binary = ifelse(cnum > 0, "Yes", "No"),
         travelTime = round(travelTime/60, 2)) %>% 
  select(travelTime, cumDrive, CT_binary) %>% 
  ggplot(aes(cumDrive, travelTime)) + 
  geom_point(alpha = 0.2) + 
  xlab("Cumulative driving time in a shift (hours)") + 
  ylab("length of time in a trip (hours)")+
  ggtitle("Trip time over cumulative driving time among CT trips")+ 
  geom_smooth(span = 0.3, size = 2)+
  facet_grid(.~CT_binary)
```

```{r}
t %>% 
  mutate(CT_binary = ifelse(cnum > 0, "Yes", "No"),
         travelTime = round(travelTime/60, 2)) %>% 
  select(travelTime, cumDrive, CT_binary) %>% 
  ggplot(aes(cumDrive, travelTime, color = CT_binary)) + 
  geom_point(alpha = 0.05) + 
  xlab("Cumulative driving time in a shift (hours)") + 
  ylab("length of time in a trip (hours)")+
  ggtitle("Trip time over cumulative driving time among CT trips")+ 
  geom_smooth()
```


## Caculating cumulative driving time by adding up CT in each trip

```{r}
pacman::p_load(tidyverse, lubridate, tidyr)
t = readRDS("data/t.Rds")

denom = t %>% 
  group_by(driver_shift_id) %>% 
  summarise(shifttime = max(cumDrive)) %>% 
  mutate(hos_int = floor(shifttime), #hour of the shift, hos
         hos_rem = shifttime - hos_int) 

denvect = vector(length = 12)
for (i in 1:12){
  denvect[i] = sum(denom$hos_rem[denom$hos_int == (i-1)]) + sum(denom$hos_int >= i)
}


z = data.frame(x = 1:10, group = c(rep(1, 5), rep(2, 5)))
z %>% group_by(group) %>% 
  mutate(x = lag(x, default = 0))

t1 = t %>% group_by(driver_shift_id) %>% 
  mutate(cumDrivelag = lag(cumDrive, default = 0)) %>% 
  ungroup()

CE_tripmax = max(t$cnum[t$cnum >0 & t$cnum<50])
CEDat = t1 %>% 
  filter(cnum > 0 & cnum < 50) %>% 
  separate(DATIME, sep = ";", into = paste0("CE_datime", 1:CE_tripmax)) %>% 
  select(cumDrive, starts_with("CE_datime"))


CEDat1 = CEDat %>% 
  gather(key = "nCE", value = "CEtime", -cumDrive) %>% 
  na.omit() %>% 
  mutate(CE_hour = ceiling(cumDrive)) 

nomin = CEDat1 %>% 
  group_by(CE_hour) %>% 
  summarise(n_CE = n())

plot_perc = nomin %>% 
  mutate(n_trips = denvect,
         perc = round(n_CE*100/n_trips, 3))

plot_perc %>% 
  ggplot(aes(x = CE_hour, y = perc)) + 
  geom_bar(stat = 'identity', fill = "#4271AE") + theme_bw() +
  xlab("Hour of the shift") + ylab("Critical event rates (%)") +
  ggtitle("The rate (%) of critical events in each hour of the shift")
```

## Calculating CT by adding driving time in each previous trip and time to CE in the last trip

```{r}
pacman::p_load(tidyverse, lubridate, tidyr)
t = readRDS("data/t.Rds")

denom = t %>% 
  group_by(driver_shift_id) %>% 
  summarise(shifttime = max(cumDrive)) %>% 
  mutate(hos_int = floor(shifttime), #hour of the shift, hos
         hos_rem = shifttime - hos_int) 

denvect = vector(length = 12)
for (i in 1:12){
  denvect[i] = sum(denom$hos_rem[denom$hos_int == (i-1)]) + sum(denom$hos_int >= i)
}


z = data.frame(x = 1:10, group = c(rep(1, 5), rep(2, 5)))
z %>% group_by(group) %>% 
  mutate(x = lag(x, default = 0))

t1 = t %>% group_by(driver_shift_id) %>% 
  mutate(cumDrivelag = lag(cumDrive, default = 0)) %>% 
  ungroup()

CE_tripmax = max(t$cnum[t$cnum >0 & t$cnum<50])
CEDat = t1 %>% 
  filter(cnum > 0 & cnum < 50) %>% 
  separate(DATIME, sep = ";", into = paste0("CE_datime", 1:CE_tripmax)) %>% 
  select(begDate, starts_with("CE_datime"), cumDrivelag)


CEDat1 = CEDat %>% 
  gather(key = "nCE", value = "CEtime", -begDate, -cumDrivelag) %>% 
  na.omit() %>% 
  mutate(CEtime = lubridate::ymd_hms(CEtime),
         withintripdiff = as.numeric(difftime(CEtime, begDate, units = "hours")),
         CEcumdrive = cumDrivelag + withintripdiff,
         CE_hour = ceiling(CEcumdrive),
         CE_hour = ifelse(CE_hour == 0, 1, CE_hour)) 

nomin = CEDat1 %>% 
  group_by(CE_hour) %>% 
  summarise(n_CE = n())

plot_perc = nomin %>% 
  mutate(n_trips = denvect,
         perc = round(n_CE*100/n_trips, 3))

plot_perc %>% 
  ggplot(aes(x = CE_hour, y = perc)) + 
  geom_bar(stat = 'identity', fill = "#4271AE") + theme_bw() +
  xlab("Hour of the shift") + ylab("Critical event rates (%)") +
  ggtitle("The rate (%) of critical events in each hour of the shift")
```

```{r eval=FALSE}
t %>% 
  dplyr::group_by(driver_shift_id) %>% 
  dplyr::arrange(driver1, begDate) %>% 
  dplyr::select(dplyr::row_number()==dplyr::n()) %>% 
  ungroup() %>% 
  ggplot(aes(x = travelTime/60)) + geom_histogram()
```


