---
title: "Rigdon Basu (1989) Time truncated data example"
author: "Miao Cai <miao.cai@slu.edu>"
date: "1/29/2019"
output:
  pdf_document:
    number_sections: true
  html_document:
    df_print: paged
linkcolor: blue
subtitle: replication in Stan
link-citations: yes
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, cache = TRUE)
```

# Time truncated data - 115 kV Transmission line example

This data is presented by @martz1975pooling. It gives the failure times, or interrpution times, of the 115 kV transmission circuit from Cunningham Generating Station, located near Hobbs, New Mexico, to Eddy County Interchange, located near Artesia, New Mexico. We assume that data collection was terminated on December 31, 1971.

Data collected in this manner, with testing terminated at a predetermined time, are called time truncated. It is important to distinguish between these approches to data collection because statistical inference precedures are different for the two situations.

```{r}
library(rstan)

t = c(0.129, 0.151, 0.762, 0.869, 2.937, 3.077, 3.841, 3.964, 4.802, 4.898, 7.868, 8.430)
trunc_time = 8.463

datstan = list(
  n = length(t),
  tau = trunc_time,
  t = t)
```

MLE estimates of $\beta$ and $theta$ are provided by @rigdon1989power:

\[
\hat{\beta} = \frac{12}{\sum_{i=1}^{12}\ln(8.463/t_i)} = 0.678
\]
$$\hat{\theta} = \frac{8.463}{12^{1/0.678}} = 0.217$$


## Diffuse priors for beta and theta - GAMMMA(1, 1)

The log likelihood function $l$ for time truncated non-homogeneous Poisson process is:

\begin{align*}
l & = \log \Big(\prod_{i=1}^n\frac{\beta}{\theta}(\frac{t_i}{\theta})^{\beta - 1}\Big)e^{-(\tau/\theta)^\beta}\\
& = \sum_{i=1}^n\log\Big(\frac{\beta}{\theta}(\frac{t_i}{\theta})^{\beta - 1}\Big) - (\frac{\tau}{\theta})^\beta\\
& = n\log\beta - n\beta\log\theta + (\beta - 1)\sum_{i=1}^n\log t_i - (\frac{\tau}{\theta})^\beta
\end{align*}

```{r}
plpstan11 = '
functions{
  real nhpp_log(vector t, real beta, real theta, real tau){
    vector[num_elements(t)] loglik_part;
    real loglikelihood;
    for (i in 1:num_elements(t)){
      loglik_part[i] = log(beta) - beta*log(theta) + (beta - 1)*log(t[i]);
    }
    loglikelihood = sum(loglik_part) - (tau/theta)^beta;
    return loglikelihood;
  }
}
data {
  int<lower=0> n; //total # of obs
  real<lower=0> tau;//truncated time
  vector<lower=0>[n] t; //failure time
}
parameters{
  real<lower=0> beta;
  real<lower=0> theta;
}
model{
  t ~ nhpp(beta, theta, tau);
//PRIORS
  beta ~ gamma(1, 1);
  theta ~ gamma(1, 1);
}
'

fitplp <- stan(
  model_code=plpstan11, model_name="NHPP", data=datstan, 
  iter=5000,warmup = 2000, chains=1, seed = 123
)

fitplp
```

Compared to MLE estimate of $\hat{\beta} = 0.678, \hat{\theta} = 0.217$.

## Weakly informative priors for beta and theta - GAMMMA(1, 5)

```{r}
plpstan15 = '
functions{
  real nhpp_log(vector t, real beta, real theta, real tau){
    vector[num_elements(t)] loglik_part;
    real loglikelihood;
    for (i in 1:num_elements(t)){
      loglik_part[i] = log(beta) - beta*log(theta) + (beta - 1)*log(t[i]);
    }
    loglikelihood = sum(loglik_part) - (tau/theta)^beta;
    return loglikelihood;
  }
}
data {
  int<lower=0> n; //total # of obs
  real<lower=0> tau;//truncated time
  vector<lower=0>[n] t; //failure time
}
parameters{
  real<lower=0> beta;
  real<lower=0> theta;
}
model{
  t ~ nhpp(beta, theta, tau);
//PRIORS
beta ~ gamma(1, 5);
theta ~ gamma(1, 5);
}
'

fitplp15 <- stan(
  model_code=plpstan15, model_name="NHPP15", data=datstan, 
  iter=5000,warmup = 2000, chains=1, seed = 123
)


fitplp15
```

Compared to MLE estimate of $\hat{\beta} = 0.678, \hat{\theta} = 0.217$.

## Another example

This software failure times in seconds example was presented in @bar1992bayesian.

```{r}
library(rstan)

software_failure = c(
  115, 115, 198, 376, 570, 706, 1783, 1798, 1813, 1905, 1955, 2026, 2632, 
  3821, 3861, 4649, 4871, 4943, 5558, 6147, 6162, 6552, 8415, 9752, 14260, 
  15094, 18494, 18500, 23061, 26229, 36800, 37363, 40133, 40785, 46378, 
  58074, 64798, 67344)
software_fTrun = software_failure[length(software_failure)]

datsoftware = list(
  n = length(software_failure),
  tau = software_fTrun,
  t = software_failure)
```


```{r}
plpsoftware <- stan(
  model_code=plpstan11, model_name="NHPPfailureT", data=datsoftware, 
  iter=5000,warmup = 2000, chains=1, seed = 123
)

plpsoftware
```




# Failure truncation data (INCORRECT)

This failure truncation example is also from @rigdon1989power.

MLE estimates for $\beta$ and $\theta$ are:

$$\hat{\beta} = \frac{13}{\sum_{i=1}^{13-1}\ln(4596/t_i)} = 0.569$$
$$\hat{\theta} = \frac{4596}{13^{1/0.596}} = 50.7$$

```{r}
library(rstan)

t1 = c(55, 166, 205, 341, 488, 567, 731, 1308, 2050, 2453, 3115, 4017, 4596)
trunc_timet1 = t1[length(t1)]

datfailure = list(
  n = length(t1),
  tau = trunc_timet1,
  t = t1)
```

## Gamma(100, 1) as the prior

Gamma distribution parameterization in Stan:

$$\text{Gamma}(y|\alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)}y^{\alpha-1}\exp(-\beta y)$$

Here we assume that $\beta \sim gamma(1, 1), \theta \sim gamma(100, 1)$. So essentially $\beta \sim EXP(1)$, while the probability density function of $\theta$ can be re-written as:

$$f(y|\alpha, \beta) = \frac{100}{99!}y^{99}\exp(-y)$$

```{r}
PGAM = function(x, alpha, beta) return(beta^alpha/factorial(alpha)*x^(alpha - 1)*exp(-beta*x))
x = seq(1, 200, 0.1)
p = PGAM(x, 100, 1)
plot(x, p, type = "l", main = "Gamma(100, 1)")
```


```{r}
plpstan11 = '
functions{
  real nhpp_log(vector t, real beta, real theta, real tau){
    vector[num_elements(t)] loglik_part;
    real loglikelihood;
    for (i in 1:num_elements(t)){
      loglik_part[i] = log(beta) - beta*log(theta) + (beta - 1)*log(t[i]);
    }
    loglikelihood = sum(loglik_part) - (tau/theta)^beta;
    return loglikelihood;
  }
}
data {
  int<lower=0> n; //total # of obs
  real<lower=0> tau;//truncated time
  vector<lower=0>[n] t; //failure time
}
parameters{
  real<lower=0> beta;
  real<lower=0> theta;
}
model{
  t ~ nhpp(beta, theta, tau);
//PRIORS
  beta ~ gamma(1, 1);
  theta ~ gamma(100, 1);
}
'


plpfailure <- stan(
  model_code=plpstan11, model_name="NHPPfailureT", data=datfailure, 
  iter=5000,warmup = 2000, chains=1, seed = 123, init = list(chain_1 = list(beta = 0.5, theta = 100))
)

plpfailure
```

## Gamma(1, 100) as the prior

```{r}
PGAM = function(x, alpha, beta) return(beta^alpha/factorial(alpha)*x^(alpha - 1)*exp(-beta*x))
x = seq(1, 200, 0.1)
p = PGAM(x, 1, 0.01)
plot(x, p, type = "l", main = "Gamma(1, 0.01)")
```




## Rerun the model with exponential(0.01)/gamma(1, 0.01) prior for beta

Exponential distribution parameterization in Stan:

$$\text{Exponential}(y|\beta) = \beta\exp(-\beta y)$$

```{r}
PEXP = function(x, beta) return(beta*exp(-beta*x))
x = seq(1, 200, 0.1)
p = PEXP(x, 1/100)
plot(x, p, type = "l", main = "Exponential(0.01)")
```


```{r}
plpstanexp = '
functions{
  real nhpp_log(vector t, real beta, real theta, real tau){
    vector[num_elements(t)] loglik_part;
    real loglikelihood;
    for (i in 1:num_elements(t)){
      loglik_part[i] = log(beta) - beta*log(theta) + (beta - 1)*log(t[i]);
    }
    loglikelihood = sum(loglik_part) - (tau/theta)^beta;
    return loglikelihood;
  }
}
data {
  int<lower=0> n; //total # of obs
  real<lower=0> tau;//truncated time
  vector<lower=0>[n] t; //failure time
}
parameters{
  real<lower=0> beta;
  real<lower=0> theta;
}
model{
  t ~ nhpp(beta, theta, tau);
//PRIORS
  beta ~ gamma(1, 1);
  theta ~ exponential(0.01);
}
'


plpfailureexp <- stan(
  model_code=plpstanexp, model_name="NHPPfailureT", data=datfailure, 
  iter=5000,warmup = 2000, chains=1, seed = 123
  #, init = list(chain_1 = list(beta = 0.5, theta = 100))
)

plpfailureexp

plpdraws = extract(plpfailureexp)
mean(plpdraws$beta)
mean(plpdraws$theta)
```


# References