---
title: "Non-homogeneous Poisson Process (NHPP)"
author: "Miao Cai"
date: "November 28, 2018"
output:
  html_document:
    theme: readable
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
    cold_folding: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```


# Non-homogeneous Poisson Process - essential part for Stan

%{stan output.var=nhpp}%

```
data{

}
model{

}
```

```{r}
library(rstan)
fit = stan()
```


# Real data example


# Mathematical prove

## Definitions

**Intensity function** The intensity function of a point process is:

$$\lambda(t) = \lim_{\Delta t \rightarrow 0} \frac{P(N(t, t + \Delta t) \geq 1)}{\Delta t}$$


**Nonhomogeneous Poisson Process** The Nonhomogeneous Poisson Process (NHPP) is a Poisson process whose intensity function is non-constant.

When the intensity function of a NHPP has the form $\lambda(t) = (\beta / \theta)(t/\theta)^{\beta - 1}$, where $\beta > 0$ and $\theta > 0$, the process is called **power law process** (PLP).

1. **Failure truncation**: When testing stops after a predetermined number of failures, the data are said to be failure truncated.
2. **Time truncation**: Data are said to be time truncated when testing stops at a predetermined time $t$.

**Conditional probability**
\begin{align*}
P(A \cap B) & = P(A)P(B|A) \\
P(A \cap B \cap C) & = P(A)P(B|A)P(C|A \cap B)
\end{align*}

## Failure Truncated Case

### The first event

The cumulative density function (cdf) of time to the first event is $F(t_1)$: $F_1(t_1) = P(T_1 \leq t_1) = 1 - S(t_1)$.

The survival function for the first event $S_1(t_1)$ is:

\begin{align*}
S_1(t_1) & = P(T_1 > t_1) \\
 & = P(N(0, t_1) = 0) \quad N \text{is the number of events}\\
 & = e^{-\int_{0}^{t_1}\lambda_{u}du}(e^{-\int_{0}^{t_1}\lambda_{u}du})^0/0!\\
 & = e^{-\int_{0}^{t_1}\lambda_{u}du}
\end{align*}



The probability density function (pdf) of time to the first event can be calculated by taking the first order derivative of the cdf $F_1(t_1)$:

\begin{align*}
f_1(t_1) & = \frac{d}{dt_1}F_1(t_1)\\
& = \frac{d}{dt_1}[1 - S_1(t_1)] \\
& = - \frac{d}{dt_1}S_1(t_1)\\
& = - \frac{d}{dt_1}e^{-\int_{0}^{t_1}\lambda (u)du}\\
& = -(-\lambda_{t_1})e^{-\int_{0}^{t_1}\lambda (u)du}\\
& = \lambda (t_1)e^{-\int_{0}^{t_1}\lambda (u)du}
\end{align*}

If this NHPP is a PLL, we plug in the intensity function $\lambda(t) = (\beta / \theta)(t/\theta)^{\beta - 1}$, then we have:

$$f_1(t_1) = \frac{\beta}{\theta}(\frac{t_1}{\theta})^{\beta - 1}e^{-(\frac{t_1}{\theta})^\beta}, \quad t_1 > 0$$

This pdf is identical with the pdf of Weibull distribution, so we have:
$$T_1 \sim \text{Weibull}(\beta, \theta)$$


### The second event

The Survival function of the second event given the first event occurred at $t_2$ is:

\begin{align}
S_2(t_2 | t_1) & = P(T_2 > t_2 | T_1 = t)\\
& = P(N(t_1, t_2) = 0|T_1 = t_1) \\
& = e^{-\int_{t_1}^{t_2}\lambda_{u}du}[\int_{t_1}^{t_2}\lambda_{u}du]^0/0!
& = e^{-\int_{t_1}^{t_2}\lambda_{u}du}
\end{align}

The we can derive the pdf of $t_2$ conditioned on $t_1$

\begin{align}
f(t_2|t_1) & = - \frac{d}{dt_2}S_2(t_2)\\
& = - \frac{d}{dt_2}e^{-\int_{t_1}^{t_2}\lambda(u)du}\\
& = \lambda(t_2)e^{-\int_{t_1}^{t_2}\lambda(u)du}\\
& = \frac{\beta}{\theta}(\frac{t_2}{\theta})^{\beta - 1}e^{-[(\frac{t_2}{\theta})^\beta - (\frac{t_1}{\theta})^\beta]}\\
& = \frac{\frac{\beta}{\theta}(\frac{t_2}{\theta})^{\beta - 1}e^{-(\frac{t_2}{\theta})^\beta }}{e^{- (\frac{t_1}{\theta})^\beta}}, \quad t_2 > t_1
\end{align}


### all events

\begin{align}
f(t_1, t_2, \cdots, t_n) & = f(t_1)f(t_2|t_1)f(t_3|t_1, t_2) \cdots f(t_n|t_1, t_2, \dots, t_{n - 1}) \\
& = \lambda (t_1)e^{-\int_{0}^{t_1} \dot \lambda (u)du}\lambda (t_2)e^{-\int_{t_1}^{t_2} \dot \lambda (u)du}\cdots\lambda (t_n)e^{-\int_{t_{n-1}}^{t_n}\lambda (u)du}\\
& = \Big(\prod_{i=1}^n\lambda(t_i)\Big)e^{-\int_0^t\lambda(u)du}\\
& = \Big(\prod_{i=1}^n\frac{\beta}{\theta}(\frac{t_i}{\theta})^{\beta - 1}\Big)e^{-(t_n/\theta)^\beta}, \quad t_1 < t_2 < \cdots < t_n
\end{align}


## Time Truncated Case

### Conditional likelihood function

We assume that the truncated time is $\tau$. The derivation of $f(t_1, t_2, \cdots, t_n|n)$ is messy in math, we directly give the conclusion here:

$$f(t_1, t_2, \cdots, t_n|n) = n!\prod_{i=1}^n\frac{\lambda(t_i)}{\Lambda(\tau)}$$

### Joint likelihood function

Therefore, the joint likelihood function for $f(n, t_1, t_2, \cdots, t_n)$ is:

\begin{align}
f(n, t_1, t_2, \cdots, t_n) & = f(n)f(t_1, t_2, \cdots, t_n|n)\\
& = \frac{e^{-\int_0^\tau \lambda(u)du}[\int_0^\tau \lambda(u)du]^n}{n!}n!\frac{\prod_{i=1}^n\lambda(t_i)}{[\Lambda(\tau)]^n}\\
& = \Big(\prod_{i=1}^n\lambda(t_i) \Big)e^{-\int_0^\tau \lambda(u)du}\\
& = \Big(\prod_{i=1}^n\frac{\beta}{\theta}(\frac{t_i}{\theta})^{\beta - 1} \Big)e^{-(\tau/\theta)^\beta},\\ 
n & = 0, 1, 2, \cdots, \quad  0 < t_1 < t_2 < \cdots < t_n
\end{align}

### Log likelihood function

The log likelihood function $l$ is then:

\begin{align}
l & = \Big(\sum_{i=1}^n\log\frac{\beta}{\theta}(\frac{t_i}{\theta})^{\beta - 1}\Big) - (\frac{\tau}{\theta})^\beta
\end{align}
